Analysis of competitive surfing tournaments with generalized Bradley-Terry

### 2021-08-24 inst/surfing_jsa.tex
slight changes following conversation with Tim

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact six out of eight querterfinalists at 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \cite{douglas2020}.

Here, we study two aspects of this sporting phenomenon: the suggested
Brazilian preference for beach break and point break reefs; and the
noncompetitive effect of compatriot rivalry.

\subsection{Brazilian beach type prefereneces}

The statistical method described here can test many different
hypotheses drawn from surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers such as
Ferreira and Toledo tend to have skillsets that favour point and beach
break beaches, those types of beach being readily accessible to these
surfers' native Brazil; see~\cite{scarfe2003} for a scientific
overview of surf break characteristics, and~\cite{butt2004} for a
surfer's perspective.

We now assess the null that
Brazilians perform equally well on all types of beaches.  To that end,
we create a reified surfer, ``Brazilian beachtype", whose strength
helps all Brazilian surfers when surfing on their (putative) preferred
beach type, an analogue of the ``home ground monster" who helps the
home team as per equation~\ref{home_ground_advantage}.  Here we modify
the likelihood function only when a single Brazilian is in the water,
to avoid possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred beach types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
beachtype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_beachtype_support}).

\begin{figure}
  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
\caption{Profile support \label{brazilian_beachtype_support} for the
  Brazilian beachtype reified entity.  The horizontal axis specifies a
  particular strength for Brazilian beachtype, and the height of the
  curve gives the support conditional on the x-axis}
\end{figure}

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic squad would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.

Such hypotheses are difficult to test with conventional techniques as
competitors's strengths themselves are nuisance
parameters~\cite{basu1977}.

Consider, as an example, the case of Smith, Buchan, Freestone who were
simultaneously in the water at Bells Beach, Australia.  Now, Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing squad, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?  We would suggest that Smith is able to
modify his strategy to benefit from the mutual rivalry between
Buchan and Freestone; and indeed we see Smith winning that particular
heat.

Reified Bradley-Terry technique furnishes an intuitive statistical
test for such mechanisms, that automatically controls for competitors'
strengths.  Noting
that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

We can now apply the Method of Support to the hypothesis that $p_C=0$,
that is, that such noncompetitive behaviour is absent.  We find that
the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
the ``real'' surfers, but the support for a nonzero $p_C$ is about 1.8
[asymptotic $p=0.28$], slightly short of the two units of support
required for rejection of our null.  We conclude that there is
evidence, but no strong evidence, to suggest that such noncompetitive
behaviour occurred---at least not from the order statistics we
considered.

@@

We discuss and assess the ``Brazilian storm'' phenomenon using reified
entities and show that there is no strong evidence to support the
supposed Brazilian preference for point and beach break beaches.  We
also find no evidence to suggest that surfing's being included in the
2020 Summer Olympics results in modified strategies to bolster
intranational ranking at the expense WCT performance.

Further work might include  {\bf\large Tim we should discuss}

>>>>

In surfing lore, the ``Brazilian storm'' refers to the emergence of
Brazil as a dominant nation of elite surfers.  The term was first used
to describe the fact that six out of eight querterfinalists at the 2011 Nike
Lowers Pro were Brazilian; it rose in public conciousness following
the victories of Medina and de Souza in the World Surf League
championships in 2014 and 2015 respectively.  The storm shows no signs
of abating, with commentators such as Douglas opining that this
generation ``represents a seismic shift for Brazilian competitors onto
an entirely new plane'' \cite{douglas2020}.

Here, we study two aspects of this sporting phenomenon: the suggested
Brazilian preference for beach break and point break wave types; and the
noncompetitive effect of compatriot rivalry.

\subsection{Brazilian wave type prefereneces}

The statistical method described here can test many different
hypotheses inspired by surfing lore.  For example, one frequent
suggestion~\cite{burgess2020,ho2021} is that Brazilian surfers tend to
have skillsets that favour point and beach break wave types, readily
accessible to these surfers' native Brazil; see~\cite{scarfe2003} for
a scientific overview of surf break characteristics,
and~\cite{butt2004} for a surfer's perspective.

We now assess the null that Brazilians perform equally well on all
types of wave types.  To that end, we create a reified surfer,
``Brazilian wavetype", whose strength helps all Brazilian surfers
when surfing on their (putative) preferred wave type, an analogue of
the ``home ground monster" who helps the home team as per
equation~\ref{home_ground_advantage}.  Here we modify the likelihood
function only when a single Brazilian is in the water, thus avoiding
possible interaction with the compatriot issue discussed in
section~\ref{compatriot}.  The null would be that this reified entity
has zero strength.  We fail to reject this null (differential support
= 0.035; $p=0.79$) and conclude that there is no evidence to suggest
that Brazilians' performce is better on their preferred wave types
than one would expect by chance.  We may go further using the method
of support: not only is the maximum likelihood estimate for Brazilian
wavetype very small (about 0.002), but in addition we may reject the
the proposal that the strength of this entity exceeds about 0.025
(Figure~\ref{brazilian_wavetype_support}).

\begin{figure}
  \includegraphics[width=4in]{surfing_files/figure-html/plotwidep-1.png}
\caption{Profile support \label{brazilian_wavetype_support} for the
  Brazilian wavetype reified entity.  The horizontal axis specifies a
  particular strength for Brazilian wavetype, and the height of the
  curve gives the support conditional on the x-axis}
\end{figure}

@@

One measure of the rising popularity of surfing is its inclusion in
the 2020 Summer Olympics~\cite{tulloch2019}, and it is natural to ask
whether this would affect the competition in some way~\cite{ho2021}.
One suggestion would be that competitors become more focussed on
beating compatriots than surfers from other countries, the thinking
being that admission to a country's Olympic team would be of more
importance to a competitor than their overall standing in the
tournament.  One might expect competitors to value their ranking when
restricted to within their country more than their international
ranking.  If this is the case then we would expect competitors to
change their behaviour if they are in the water with another surfer of
the same nationality.  Specifically, if we have three surfers in the
water of whom two are of one nationality and the third of another, we
would expect the compatriots to adopt a style calculated to beat their
countryman, possibly at the expense of performance relative to the odd
man out.  Such hypotheses are difficult to test with conventional
techniques as competitors's strengths themselves are nuisance
parameters~\cite{basu1977}.

Consider, as an example, the case of Smith, Buchan, Freestone who were
simultaneously in the water at Bells Beach, Australia.  Now, Buchan
and Freestone are Australian, and Jordy Smith is South African.  It is
reasonable to ask whether Buchan and Freestone, who are in direct
competition with one another for a place in the Australian Olympic
surfing team, would be more motivated to beating each other than to
come ahead of Smith.  This would be a rational response, but arguably
noncompetitive.  How would such a strategy manifest itself in
scorelines?  We would suggest that Smith is able to modify his
strategy to benefit from the mutual rivalry between Buchan and
Freestone; and indeed we see Smith winning that particular heat.  We
tentatively interpret this as a Nash-type equilibrium: there is no
incentive for either Buchan nor Freestone to deviate from their
prioritization of defeating their compatriot; and indeed neither
Buchan nor Freestone were selected for the 2020 Olympic team.

The reified Bradley-Terry technique furnishes an intuitive statistical
test for such mechanisms, that automatically controls for competitors'
strengths.  Noting
that the compatriots effectively place themselves at a {\em
  disadvantage} when compared with the odd man out, we introduce a
compatriot monster whose strength $p_C$ helps the odd man out.
Writing $p_1, p_2$ for the strengths of the compatriots and $q$ for
the odd man out, we construct likelihood functions for the six
possible orderings when two compatriots and an odd man out are in the
water at the same time in Table~\ref{likelihoodcompatriot} (if we have
either three different nationalities or three identical nationalities,
we revert to the standard Plackett-Luce likelihood).  Note that all
six likelihood functions in table~\ref{likelihoodcompatriot} are
informative about $p_1,p_2,q$ in addition to being informative about
the object of inference, $p_C$.

@@

We can now apply the Method of Support to the hypothesis that $p_C=0$,
that is, that such noncompetitive behaviour is absent.  We find that
the maximum likelihood estimate for $p_C$ is about 0.015, weaker than
the ``real'' surfers, but the support for a nonzero $p_C$ is about
0.59 [asymptotic $p=0.28$], insufficient to reject our null.  We
conclude that there is no strong evidence to suggest that such
noncompetitive behaviour occurred---at least not from the order
statistics we considered.

@@

We discuss and assess the ``Brazilian storm'' phenomenon using reified
Bradley-Terry techniques, and show that there is no strong evidence to
support the supposed Brazilian preference for point and beach break
beaches.  We also find no evidence to suggest that surfing's being
included in the 2020 Summer Olympics results in modified strategies to
bolster intranational ranking at the expense WCT performance.

Further work might include statistical investigation of noncompetitive
NASH STUFF compatriot priority infringements between compatriots.
